---
title: "Joel Embiid's Impact"
author: Henry Han
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(rvest)
library(tidyr)
library(lubridate)
library(ggplot2)
library(ggthemes)
```

Joel Embiid has been a matchup nightmare ever since he debuted as a rookie last year for the Philadelphia 76ers. In just 31 games played before he had a season-ending injury, he showcased a skillset that drew comparisons to NBA legends Hakeem Olajuwon and Shaquille O'Neal. How can a rookie turn so many heads in just a handful of games? A 250lb 7-footer who played soccer and volleyball until junior year in high school, Joel is a monster of a player who can out-physical anyone in the NBA, out run any big man in the NBA, and has a shooting touch to shoot 36% from beyond the 3-point arc as a rookie. The question then is, how exactly do the 76ers get better or worse when Joel plays on the court? Just how dominating has he been? To start, we need to scrape the data from [basketball-reference](https://www.basketball-reference.com/players/e/embiijo01/gamelog/2018/).

```{r}
url <- "https://www.basketball-reference.com/players/e/embiijo01/gamelog/2018/"

# scrape html and store table
tables <- url %>%
  read_html() %>%
  html_nodes(css = "table")

stats <- html_table(tables[[8]], fill=TRUE)
stats
```

There are a couple of columns that aren't necessary in our analysis. We know that Joel Embiid plays for the 76ers (PHI) and the date and age columns aren't relevant. In addition, there are a few rows in the table marked with "inactive", "not with team", "did not dress" or "did not play", which are various reasons that Joel did not play in those games. From there, we created a table that shows all the games that Joel Embiid did not play in this season. Notice that the column GS has a value of 1 (indicating that he started every game he played) for each game that Joel did play, giving us a straightforward way to filter the rows we need. It appears that the team tends to lose when Joel isn't playing, but it would be irresponsible if we made the conclusion that Joel is the clear difference maker. Let's continue by taking a look at the 76ers offensive and defensive efficiencies when Joel is playing vs when he's not. 

```{r}
# take out date, age, team and trailing cloumns
clean_stats <- stats[-c(3:5,31)]

# add in column names
colnames(clean_stats)[3] <- "Location"
colnames(clean_stats)[5] <- "WinLoseComb"

# summarize games that Joel did not play in
inactive_stats <- clean_stats %>%
  filter(Rk != "Rk") %>%
  filter(GS != 1 ) %>%
  select(Rk, Opp, WinLoseComb)

inactive_stats

```

To continue, we filtered out rows for games that Joel did not play in, as well as repeated rows containing column names. In addition, the location column was renamed and recoded to display "Away/Home" rather than "@" and nothing. We also took the WinLoseComb column and separated it into "Win/Loss" and the numerical margin of victory. 

```{r}
# filter out rows for games that Joel did not play
# also filter out additional column name rows
clean_stats <- filter(clean_stats, GS == 1)

# clean up location and margin of victory columns
clean_stats <- clean_stats %>%
  mutate(Location = as.factor(ifelse(Location == '@', "Away", "Home"))) %>%
  mutate(WinLose = as.factor(ifelse(grepl("W", WinLoseComb), "Win", "Lose"))) %>%
  mutate(Margin = gsub("L", "", WinLoseComb)) %>%
  mutate(Margin = gsub("W", "", Margin)) %>%
  mutate(Margin = gsub("\\)", "", Margin)) %>%
  mutate(Margin = gsub("\\(", "", Margin)) %>%
  mutate(Margin = gsub(" ", "", Margin)) %>%
  mutate(Margin = as.numeric(gsub("+", "", Margin)))

clean_stats <- clean_stats[-c(5)]

# clean up miscellaneous variable names and formats
clean_stats <- clean_stats %>%
  mutate(`+/-` = gsub("+", "", `+/-`)) %>%
  mutate_each(funs(as.numeric), c(GS, FG, FGA, `FG%`, `3P`, `3PA`, `3P%`, FT, FTA, `FT%`, ORB, DRB, TRB, AST, STL, BLK, TOV, PF, PTS, GmSc, `+/-`, Margin)) %>%
  mutate_each(funs(as.factor), c(Rk, G, Location, Opp, WinLose)) %>%
  mutate(MP = round(as.numeric(ms(MP))/60, 2))
  
glimpse(clean_stats)

```

There are 55 games in total in our data set. These include 55 starts, 33 wins, 22 losses and 34 double-doubles. From there, we can create some summary tables on how Joel Embiid has performed so far this season. In addition, we can create multiple histograms that show the distributions of some of the key statistics. 

```{r}
# create summary table for key stats
df_sum <- clean_stats %>%
  select(MP, PTS, TRB, AST, STL, BLK) %>% 
  summarise_each(funs(mean = mean, sd = sd, min = min, max = max))

df_sum_tidy <- df_sum %>% 
  gather(stat, val) %>%
  separate(stat, into = c("var", "stat"), sep = "_") %>%
  spread(stat, val) %>%
  select(var, mean, sd, min, max) %>%
  mutate(mean = round(mean, 2), sd = round(sd, 2)) 

df_sum_tidy

# create histograms for key stats
plot1 <- clean_stats %>%
  select(MP, PTS, TRB, AST, STL, BLK) %>%
  gather() %>%                             
  ggplot(aes(x = value)) +                     
    facet_wrap(~ key, scales = "free") +   
    geom_histogram(fill = "coral1") + 
    labs(title = "Distributions of Key Stats for Joel Embiid", subtitle = "Data from basketball reference", x = "Value", y = "Count") +
    theme_economist()

plot1

```

To take a closer look at Joel's impact while he's on the court, we can look at his advanced stats, which can also be found at [basketball-reference]("https://www.basketball-reference.com/players/e/embiijo01/gamelog-advanced/2018/"). 

```{r}
adv_url <- "https://www.basketball-reference.com/players/e/embiijo01/gamelog-advanced/2018/"

# scrape html and store table
tables <- adv_url %>%
  read_html() %>%
  html_nodes(css = "table")

adv_stats <- html_table(tables[[1]], fill=TRUE)

```

As you can see, the advanced game log can also use some of the same initial wrangling to clean up the data. 

```{r}
# initial wrangling
clean_adv_stats <- adv_stats[-c(3:10,23)]
clean_adv_stats <- clean_adv_stats %>%
  filter(Rk != "Rk")

```

Let's join the two datasets so that it is easier to see which games the advanced stats are from. 

```{r}
# join the normal and advanced game log tables
combined_stats <- full_join(clean_stats, clean_adv_stats) %>%
  filter(!is.na(Location)) %>%
  mutate_each(funs(as.numeric), c(`TS%`, `eFG%`, `ORB%`, `DRB%`, `TRB%`,`AST%`, `STL%`, `BLK%`, `TOV%`, `USG%`, ORtg, DRtg)) %>%
  arrange(`USG%`)

glimpse(combined_stats)

```

A player's usage rate is a metric that estimates the percentage of the team's possessions a player uses while they are in the game. A 25% usage rate means that 25% of plays ended with that player either shooting the ball, turning the ball over, or getting to the free throw line. Usage rate serves as a metric that shows how involved a player is in their team's offense. Let's take a closer look at Joel's usage rate and how that affects his other stats and ultimately the outcome of the game. 

```{r}
# group all games by Joel's usage rate
usage_rate_under_30 <- filter(combined_stats, `USG%` < 30)
usage_rate_30_40 <- filter(combined_stats, `USG%` < 40 & `USG%` >= 30)
usage_rate_above_40 <- filter(combined_stats, `USG%` >= 40) 

# plot minutes played against points scored
plot2 <- ggplot(data=combined_stats, aes(x=MP, y=`PTS`, shape=WinLose)) +
  geom_point(data=usage_rate_under_30, alpha=0.6, size = 7, color='red') +
  geom_point(data=usage_rate_30_40, alpha=0.6, size = 7, color='green') + 
  geom_point(data=usage_rate_above_40, alpha=0.6, size = 7, color='blue') + 
  labs(x="Minutes Played", y="Points", title="Minutes Played vs. Points Scored", subtitle="Data from basketball reference\n\nRed: USG<30%, Green: 30%<USG<40%, Blue: 40%<USG") +
  theme_economist()

plot2

```

Initially, we can see that there is a slight upward trend between minutes played and points scored. However, there is no clear trend between minutes played and usage rate. There is a slight trend that the higher Joel's usage rate, the more points he scores. Let's go further and look at more specific stats like +/-: a measure of how the team fares against their opponent when a player is on the court. For example, a player with a +/- of 8 means that when that player is on the court, their team outscores their opponent by 8 points. 

```{r}
# summarize key stats by usage rate

# summary for games with usage rate above 40%
summary_above_40 <- usage_rate_above_40 %>%
  summarise(`USG%` = round(mean(`USG%`), 2),
            `max +/-` = round(max(`+/-`), 2),
            `min +/-` = round(min(`+/-`), 2),
            `+/-` = round(mean(`+/-`), 2),
            Margin = round(mean(Margin), 2),
            numWins = sum(WinLose == 'Win'),
            numLosses = sum(WinLose == 'Lose'),
            MP = round(mean(MP), 2),
            PTS = round(mean(PTS), 2),
            TRB = round(mean(TRB), 2),
            AST = round(mean(AST), 2),
            STL = round(mean(STL), 2),
            BLK = round(mean(BLK), 2),
            TOV = round(mean(TOV), 2),
            `eFG%` = round(mean(`eFG%`), 2),
            `TS%` = round(mean(`TS%`), 2))

# summary for games with usage rate under 30%
summary_under_30 <- usage_rate_under_30 %>%
  summarise(`USG%` = round(mean(`USG%`), 2),
            `max +/-` = round(max(`+/-`), 2),
            `min +/-` = round(min(`+/-`), 2),
            `+/-` = round(mean(`+/-`), 2),
            Margin = round(mean(Margin), 2),
            numWins = sum(WinLose == 'Win'),
            numLosses = sum(WinLose == 'Lose'),
            MP = round(mean(MP), 2),
            PTS = round(mean(PTS), 2),
            TRB = round(mean(TRB), 2),
            AST = round(mean(AST), 2),
            STL = round(mean(STL), 2),
            BLK = round(mean(BLK), 2),
            TOV = round(mean(TOV), 2),
            `eFG%` = round(mean(`eFG%`), 2),
            `TS%` = round(mean(`TS%`), 2))

# summary for games with usage rate between 30% and 40%
# first, impute to account for a "NA" in the `+/-` column
ds_preproc <- preProcess(usage_rate_30_40, method = c("medianImpute"))
usage_rate_30_40_proc <- predict(ds_preproc, usage_rate_30_40)

summary_30_40 <- usage_rate_30_40_proc %>%
  summarise(`USG%` = round(mean(`USG%`), 2),
            `max +/-` = round(max(`+/-`), 2),
            `min +/-` = round(min(`+/-`), 2),
            `+/-` = round(mean(`+/-`), 2),
            Margin = round(mean(Margin), 2),
            numWins = sum(WinLose == 'Win'),
            numLosses = sum(WinLose == 'Lose'),
            MP = round(mean(MP), 2),
            PTS = round(mean(PTS), 2),
            TRB = round(mean(TRB), 2),
            AST = round(mean(AST), 2),
            STL = round(mean(STL), 2),
            BLK = round(mean(BLK), 2),
            TOV = round(mean(TOV), 2),
            `eFG%` = round(mean(`eFG%`), 2),
            `TS%` = round(mean(`TS%`), 2))

summary_combined <- bind_rows(summary_under_30, summary_30_40, summary_above_40)
summary_combined <- cbind(`Usage Rate` = c('20-30%', '30-40%', '40%+'), summary_combined)

summary_combined

```

Although the sample sizes are small, it appears that Joel seems to perform the best when his usage rate is below 30% and above 40%. Regardless of usage rate, Joel averaged 52% efective field goal percentage and 57% true shooting percentage. This is interesting because Joel's scoring output is drastically different depending on his usage rate: 27.33 points per game when his usage rate is above 40% and 22 points per game when it's below 30%. When Joel's usage rate is above 40%, the 76ers won a whopping 5 out of 6 times by an average margin of 5.33 points. His average scoring was the highest in this usage range, but also averaged the most turnovers. It appears that when Joel's usage rate is this high, he also appears to contribute less in other statistical categories - he averaged the least amount of assists, steals, and blocks when his usage rate exceeded 40%. On the contrary, Joel averaged the most minutes and least amount of points when his usage rate was below 30% - though the team won a healthy 5 out of 7 games. It appears that when Joel's usage rate is below 30%, he is contributing a lot in other ways - he averaged the most assists, steals, and blocks, and the least amount of turnovers. 

When Joel's usage rate is between 30 and 40%, his impact on the game is less distinct. The team essentially went 0.500 (22-19) and Joel averaged the least amount of minutes and rebounds. Although his stats aren't the best in this usage rate range, his plus/minus was actually the highest at 6.95. This means that when Joel is on the floor, the team is outscoring their opponents by an average of 6.95 points. Perhaps he should play more minutes if his ussage rate is in this range? 

A bigger sample size is necessary to make more concrete conclusions, but some observations can still be made. It appears that when the game is run through Joel the most, when Joel's usage rate is above 40%, the 76ers tend to win more (6-1) and by more as well. On the other end, when the game isn't run through Joel as much, when Joel's usage rate is below 30%, the 76ers still tend to win (5-2). When Joel's usage rate is below 30%, he actually averages the most minutes, assists, and contributes the most on defense as well. If conclusions were to be made in such a small sample size, either let Joel take over the offense (high usage rate), or let Joel play more minutes and take over on defense (low usage rate). 
