---
title: "Joel Embiid's Impact"
author: Henry Han
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(rvest)
library(tidyr)
library(lubridate)
library(ggplot2)
library(ggthemes)
```

Joel Embiid has been a matchup nightmare ever since he debuted as a rookie last year for the Philadelphia 76ers. In just 31 games played before he had a season-ending injury, he showcased a skillset that drew comparisons to NBA legends Hakeem Olajuwon and Shaquille O'Neal. How can a rookie turn so many heads in just a handful of games? A 250lb 7-footer who played soccer and volleyball until junior year in high school, Joel is a monster of a player who can out-physical anyone in the NBA, out run any big man in the NBA, and has a shooting touch to shoot 36% from beyond the 3-point arc as a rookie. The question then is, how exactly do the 76ers get better or worse when Joel plays on the court? Just how dominating has he been? To start, we need to scrape the data from [basketball-reference](https://www.basketball-reference.com/players/e/embiijo01/gamelog/2018/).

```{r}
url <- "https://www.basketball-reference.com/players/e/embiijo01/gamelog/2018/"

# scrape html and store table
tables <- url %>%
  read_html() %>%
  html_nodes(css = "table")

stats <- html_table(tables[[8]], fill=TRUE)
stats
```

There are a couple of columns that aren't necessary in our analysis. We know that Joel Embiid plays for the 76ers (PHI) and the date and age columns aren't relevant. In addition, there are a few rows in the table marked with "inactive", "not with team", "did not dress" or "did not play", which are various reasons that Joel did not play in those games. From there, we created a table that shows all the games that Joel Embiid did not play in this season. Notice that the column GS has a value of 1 (indicating that he started every game he played) for each game that Joel did play, giving us a straightforward way to filter the rows we need. It appears that the team tends to lose when Joel isn't playing, but it would be irresponsible if we made the conclusion that Joel is the clear difference maker. Let's continue by taking a look at the 76ers offensive and defensive efficiencies when Joel is playing vs when he's not. 

```{r}
# take out date, age, team and trailing cloumns
clean_stats <- stats[-c(3:5,31)]

# add in column names
colnames(clean_stats)[3] <- "Location"
colnames(clean_stats)[5] <- "WinLoseComb"

# summarize games that Joel did not play in
inactive_stats <- clean_stats %>%
  filter(Rk != "Rk") %>%
  filter(GS != 1 ) %>%
  select(Rk, Opp, WinLoseComb)

inactive_stats

```

To continue, we filtered out rows for games that Joel did not play in, as well as repeated rows containing column names. In addition, the location column was renamed and recoded to display "Away/Home" rather than "@" and nothing. We also took the WinLoseComb column and separated it into "Win/Loss" and the numerical margin of victory. 

```{r}
# filter out rows for games that Joel did not play
# also filter out additional column name rows
clean_stats <- filter(clean_stats, GS == 1)

# clean up location and margin of victory columns
clean_stats <- clean_stats %>%
  mutate(Location = as.factor(ifelse(Location == '@', "Away", "Home"))) %>%
  mutate(WinLose = as.factor(ifelse(grepl("W", WinLoseComb), "Win", "Lose"))) %>%
  mutate(Margin = gsub("L", "", WinLoseComb)) %>%
  mutate(Margin = gsub("W", "", Margin)) %>%
  mutate(Margin = gsub("\\)", "", Margin)) %>%
  mutate(Margin = gsub("\\(", "", Margin)) %>%
  mutate(Margin = gsub(" ", "", Margin)) %>%
  mutate(Margin = as.numeric(gsub("+", "", Margin)))

clean_stats <- clean_stats[-c(5)]

# clean up miscellaneous variable names and formats
clean_stats <- clean_stats %>%
  mutate(`+/-` = gsub("+", "", `+/-`)) %>%
  mutate_each(funs(as.numeric), c(GS, FG, FGA, `FG%`, `3P`, `3PA`, `3P%`, FT, FTA, `FT%`, ORB, DRB, TRB, AST, STL, BLK, TOV, PF, PTS, GmSc, `+/-`, Margin)) %>%
  mutate_each(funs(as.factor), c(Rk, G, Location, Opp, WinLose)) %>%
  mutate(MP = round(as.numeric(ms(MP))/60, 2))
  
glimpse(clean_stats)

```

There are 55 games in total in our data set. These include 55 starts, 33 wins, 22 losses and 34 double-doubles. From there, we can create some summary tables on how Joel Embiid has performed so far this season. In addition, we can create multiple histograms that show the distributions of some of the key statistics. 

```{r}
# create summary table for key stats
df_sum <- clean_stats %>%
  select(MP, PTS, TRB, AST, STL, BLK) %>% 
  summarise_each(funs(mean = mean, sd = sd, min = min, max = max))

df_sum_tidy <- df_sum %>% 
  gather(stat, val) %>%
  separate(stat, into = c("var", "stat"), sep = "_") %>%
  spread(stat, val) %>%
  select(var, mean, sd, min, max) %>%
  mutate(mean = round(mean, 2), sd = round(sd, 2)) 

df_sum_tidy

# create histograms for key stats
plot1 <- clean_stats %>%
  select(MP, PTS, TRB, AST, STL, BLK) %>%
  gather() %>%                             
  ggplot(aes(x = value)) +                     
    facet_wrap(~ key, scales = "free") +   
    geom_histogram(fill = "coral1") + 
    labs(title = "Distributions of Key Stats for Joel Embiid", subtitle = "Data from basketball reference", x = "Value", y = "Count") +
    theme_economist()

plot1

```

To take a closer look at Joel's impact while he's on the court, we can look at his advanced stats, which can also be found at [basketball-reference]("https://www.basketball-reference.com/players/e/embiijo01/gamelog-advanced/2018/"). 

```{r}
adv_url <- "https://www.basketball-reference.com/players/e/embiijo01/gamelog-advanced/2018/"

# scrape html and store table
tables <- adv_url %>%
  read_html() %>%
  html_nodes(css = "table")

adv_stats <- html_table(tables[[1]], fill=TRUE)

```

As you can see, the advanced game log can also use some of the same initial wrangling to clean up the data. 

```{r}
# initial wrangling
clean_adv_stats <- adv_stats[-c(3:10,23)]
clean_adv_stats <- clean_adv_stats %>%
  filter(Rk != "Rk")

```

Let's join the two datasets so that it is easier to see which games the advanced stats are from. 

```{r}
# join the normal and advanced game log tables
combined_stats <- full_join(clean_stats, clean_adv_stats) %>%
  filter(!is.na(Location)) %>%
  mutate_each(funs(as.numeric), c(`TS%`, `eFG%`, `ORB%`, `DRB%`, `TRB%`,`AST%`, `STL%`, `BLK%`, `TOV%`, `USG%`, ORtg, DRtg)) %>%
  arrange(`USG%`)

glimpse(combined_stats)

```

A player's usage rate is a metric that estimates the percentage of the team's possessions a player uses while they are in the game. A 25% usage rate means that 25% of plays ended with that player either shooting the ball, turning the ball over, or getting to the free throw line. Usage rate serves as a metric that shows how involved a player is in their team's offense. Let's take a closer look at Joel's usage rate and how that affects his other stats and ultimately the outcome of the game. 

```{r}
# group all games by Joel's usage rate
usage_rate_under_30 <- filter(combined_stats, `USG%` < 30)
usage_rate_30_40 <- filter(combined_stats, `USG%` < 40 & `USG%` >= 30)
usage_rate_above_40 <- filter(combined_stats, `USG%` >= 40) 

# plot minutes played against points scored
plot2 <- ggplot(data=combined_stats, aes(x=MP, y=`PTS`, shape=WinLose)) +
  geom_point(data=usage_rate_under_30, alpha=0.6, size = 7, color='red') +
  geom_point(data=usage_rate_30_40, alpha=0.6, size = 7, color='green') + 
  geom_point(data=usage_rate_above_40, alpha=0.6, size = 7, color='blue') + 
  labs(x="Minutes Played", y="Points", title="Minutes Played vs. Points Scored", subtitle="Data from basketball reference\n\nRed: USG<30%, Green: 30%<USG<40%, Blue: 40%<USG") +
  theme_economist()

plot2

```
